---
title: "R Notebook"
output: html_notebook
---


# Initial Analysis

## Load libraries

```{r}
library('ggplot2') # visualisation
library('scales') # visualisation
library('grid') # visualisation
library('RColorBrewer') # visualisation
library('alluvial') # visualisation
library('dplyr') # data manipulation
library('readr') # input/output
library('data.table') # data manipulation
library('tibble') # data wrangling
library('tidyr') # data wrangling
library('stringr') # string manipulation
library('forcats') # factor manipulation
library('lubridate') # date and time
library('geosphere') # geospatial locations
library('leaflet') # maps
library('leaflet.extras') # maps
library('maps') # maps

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

## Some parameter settings

```{r}
SEED=20190216
```

## Load data

```{r}
train_dt = fread("~/GitHub/TripDuration/data/raw/train_sub.csv")
test_dt = fread("~/GitHub/TripDuration/data/raw/test.csv")
```

```{r}
# data from 02_OSRM.Rmd
#fast_route = readRDS("../data/interim/fast_route_basic.RDS")
```


```{r}
c(nrow(train_dt), nrow(test_dt))
head(train_dt)
head(test_dt)
```

`start_timestamp` is an unix time.


## Missing Values

```{r}
sum(is.na(train_dt))
sum(is.na(test_dt))
```

## Combining train and test

```{r}
cb_dt = bind_rows(train_dt %>% mutate(dset = "train"),
                   test_dt %>% mutate(dset = "test",
                                     duration = NA))

cb_dt = cb_dt %>% mutate(dset = factor(dset))
```

### Time range for train and test

```{r}
ggplot(cb_dt %>% sample_n(10000), aes(x = start_timestamp, y = dset)) + geom_point() + theme_bw(base_size = 15)
```

It seems that the time range is overlapped between training and testing set.

### Location range for train and test

```{r}
set.seed(SEED)

tmp = cb_dt %>% group_by(dset) %>% sample_n(5000)

tmp %>%
  filter(start_lng > -75 & start_lng < -73) %>%
  filter(start_lat > 40 & start_lat < 42) %>%
  ggplot(aes(x = start_lng, y = start_lat, color = dset)) +
  geom_point(alpha = 0.1, size = 0.2) +
  coord_cartesian(xlim = c(-74.02,-73.77), ylim = c(40.63,40.84)) +
  facet_wrap(~dset)

tmp %>%
  filter(dset == "train") %>%
  leaflet %>%
  addProviderTiles("Esri.NatGeoWorldMap") %>%
  addCircleMarkers(~ start_lng, ~start_lat, radius = 1,
                   color = "blue", fillOpacity = 0.3)

tmp %>%
  filter(dset == "test") %>%
  leaflet %>%
  addProviderTiles("Esri.NatGeoWorldMap") %>%
  addCircleMarkers(~ start_lng, ~start_lat, radius = 1,
                   color = "red", fillOpacity = 0.3)
```

The location mostly focus on Manhattan area.  Another notable hot-spot is JFK and Laguadia airport.



## Feature transform

I think I can safely assume the origin date for UNIX time is 1970-01-01. However, the time zone can be tricky. I assume the number of taxi trips during the night are usually less than those during the day. The current time zone seems to make sense.

```{r}
train_dt = train_dt %>%
  mutate(pickup_datetime = with_tz(
    as_datetime(
      as.POSIXct(start_timestamp, origin="1970-01-01")
      ))
  ) %>%
  mutate(hour = hour(pickup_datetime),
         month = as.integer(month(pickup_datetime)),
         wday0 =  wday(pickup_datetime, label = TRUE),
         wday = as.integer(fct_relevel(wday0, c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri","Sat"))),
         wday_r = as.integer(fct_relevel(wday0, c("Sun", "Sat", "Mon", "Tue", "Wed", "Thu", "Fri")))
         )
```

## Feature Selection

### Distribution of trip duration
```{r}
train_dt %>%
  ggplot(aes(duration)) +
  geom_histogram(fill = "red", bins = 150) +
  scale_x_log10() +
  scale_y_sqrt()
```


### Relation between trip duration and distance

```{r}
pick_coord <- train_dt %>%

  select(start_lng, start_lat)

drop_coord <- train_dt %>%

  select(end_lng, end_lat)

train_dt$dist <- distCosine(pick_coord, drop_coord)

train_dt$bearing = bearing(pick_coord, drop_coord)

set.seed(4321)

train_dt %>%

  sample_n(5e4) %>%

  ggplot(aes(dist, duration)) +

  geom_point() +

  scale_x_log10() +

  scale_y_log10() +

  labs(x = "Direct distance [m]", y = "Trip duration [s]")


```

```{r}
train_dt %>%

  filter(duration < 3600 & duration > 120) %>%

  filter(dist > 100 & dist < 100e3) %>%

  ggplot(aes(dist, duration)) +

  geom_bin2d(bins = c(500,500)) +

  scale_x_log10() +

  scale_y_log10() +

  labs(x = "Direct distance [m]", y = "Trip duration [s]")

```



```{r}
train_dt = train_dt %>% mutate(speed=(dist/duration)*3.6)

```
### Distribution of average speed
```{r}
  train_dt %>% filter(speed > 2 & speed < 1e2)  %>%

  ggplot(aes(speed)) +

  geom_histogram(fill = "red", bins = 50) +

  labs(x = "Average speed [km/h] (direct distance)")
```

## Feature Engineering

### Time-related Features
How time-related factors (hour of day, day of week, and month) affect the trip duration. They might affect trip distance and average speed.

```{r}
for (col in c( "hour", "wday","month")) {

  p1 = train_dt %>%
  group_by_(col) %>%
  summarise(median_duration = median(duration)) %>%
  ggplot(aes_string(col, "median_duration")) +
  geom_point(size = 4) +
  labs( y = "Median Duration [s]") +
  theme(legend.position = "none")

  p2 <- train_dt %>%
  group_by_(col) %>%
  count() %>%
  ggplot(aes_string(col, "n")) +
  geom_point(size = 4) +
  labs( y = "Total number of pickups") +
  theme(legend.position = "none")

  p3 <- train_dt %>%
  group_by_(col) %>%
  summarise(median_speed = median(speed)) %>%
  ggplot(aes_string(col, "median_speed")) +
  geom_point(size = 4) +
  labs( y = "Median Speed[km/h]") +
  theme(legend.position = "none")

  p4 <- train_dt %>%
  group_by_(col) %>%
  summarise(median_dist = median(dist)) %>%
  ggplot(aes_string(col, "median_dist")) +
  geom_point(size = 4) +
  labs( y = "Median Distance [m]") +
  theme(legend.position = "none")


  layout <- matrix(c(1,2,3,4),2,2,byrow=TRUE)
  multiplot(p1, p2, p3, p4, layout=layout)

}


```

### Geo-related Features

How the geo features would affect the trip duration. How pick-up/drop-off location affect the trip duration and speed.
```{r}
p1 = train_dt %>% filter(duration > 0) %>% sample_n(400) %>% ggplot(aes(y = start_lat, x = start_lng, color = duration)) + geom_point() + scale_color_gradient(trans = "log", low = "blue", high = "red") + theme(legend.position = c( 0.8, 0.3))

p2 = train_dt %>% filter(speed > 0) %>% sample_n(400) %>% ggplot(aes(y = start_lat, x = start_lng, color = speed)) + geom_point() + scale_color_gradient(trans = "log", low = "blue", high = "red")+ theme(legend.position = c( 0.8, 0.3))

  layout <- matrix(c(1,2),1,2,byrow=TRUE)
  multiplot(p1, p2,layout=layout)

```


How bearing affects the trip duration and speed.
```{r}
p1 <- train_dt %>%
  filter(duration < 3600*22) %>%
  filter(dist < 1e5) %>%
  ggplot(aes(bearing, duration)) +
  geom_bin2d(bins = c(100,100)) +
  scale_y_log10() +
  labs(x = "Bearing", y = "Trip duration") +
  coord_polar() +
  scale_x_continuous(breaks = seq(-180, 180, by = 45))

p2 <- train_dt %>%
  filter(speed < 75 & dist < 1e5) %>%
  ggplot(aes(bearing, speed)) +
  geom_bin2d(bins = c(100,100)) +
  labs(x = "Bearing", y = "Speed") +
  coord_polar() +
  scale_x_continuous(breaks = seq(-180, 180, by = 45))


layout <- matrix(c(1,2),1,2,byrow=TRUE)
multiplot(p1, p2, layout=layout)

```


How distance to airport affects the trip duration and speed.
```{r}
jfk_coord <- tibble(lon = -73.778889, lat = 40.639722)
la_guardia_coord <- tibble(lon = -73.872611, lat = 40.77725)

train_dt$jfk_dist_pick <- distCosine(pick_coord, jfk_coord)
train_dt$jfk_dist_drop <- distCosine(drop_coord, jfk_coord)
train_dt$lg_dist_pick <- distCosine(pick_coord, la_guardia_coord)
train_dt$lg_dist_drop <- distCosine(drop_coord, la_guardia_coord)

```

```{r}

for (col in c( "jfk_dist_pick", "jfk_dist_drop","lg_dist_pick", "lg_dist_drop"))
{

  p1 = train_dt %>%
  sample_n(500) %>%
  ggplot(aes_string(col, "duration")) +
  geom_point() +
  labs( y = "Trip duration [s]") +
  theme(legend.position = "none")  

  p2 = train_dt %>%
  sample_n(500) %>%
  ggplot(aes_string(col, "speed")) +
  geom_point() +
  labs( y = "Speed [km/h]") +
  theme(legend.position = "none")

  layout <- matrix(c(1,2),1,2,byrow=TRUE)
  multiplot(p1, p2, layout=layout)

}
```
## Outlier

I have exclude data points whose trip duration 10 seconds or longer than a day.


# Final Feature Engineering

```{r}
train_dt = fread("../data/raw/train.csv")
test_dt = fread("../data/raw/test.csv")

# airport location
jfk_coord = tibble(lon = -73.778889, lat = 40.639722)
la_guardia_coord = tibble(lon = -73.872611, lat = 40.77725)


cb_dt = bind_rows(train_dt %>% mutate(dset = "train"),
                   test_dt %>% mutate(dset = "test",
                                     duration = NA))
cb_dt = cb_dt %>% mutate(dset = factor(dset))
```
## Distance related

```{r}
pick_coord <- cb_dt %>%
  select(start_lng, start_lat)
drop_coord <- cb_dt %>%
  select(end_lng, end_lat)

cb_dt$dist = distCosine(pick_coord, drop_coord)
cb_dt$bearing = bearing(pick_coord, drop_coord)

cb_dt$jfk_pick <- distCosine(pick_coord, jfk_coord)
cb_dt$jfk_drop <- distCosine(drop_coord, jfk_coord)
cb_dt$lg_pick <- distCosine(pick_coord, la_guardia_coord)
cb_dt$lg_drop <- distCosine(drop_coord, la_guardia_coord)

cb_dt <- cb_dt %>%
  mutate(speed = dist/duration*3.6,
         jfk_trip = as.integer((jfk_pick < 2e3) | (jfk_drop < 2e3)),
         lg_trip = as.integer((lg_pick < 2e3) | (lg_drop < 2e3)),
         center_latitude = (start_lat + end_lat)/2,
         center_longitude = (start_lng + end_lng)/2
         ) %>%
  rename(pickup_longitude = start_lng, pickup_latitude = start_lat,
         dropoff_longitude = end_lng, dropoff_latitude = end_lat)
```
## Time related

```{r}
cb_dt = cb_dt %>%
  mutate(pickup_datetime = with_tz(
    as_datetime(
      as.POSIXct(start_timestamp, origin="1970-01-01")
      ))
  ) %>%
  mutate(date = date(pickup_datetime),
         month = month(pickup_datetime),
         wday = wday(pickup_datetime, label = TRUE),
         wday = as.integer(fct_relevel(wday, c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))),
         hour = hour(pickup_datetime),
         minute = minute(pickup_datetime),
         whour = wday*24 + hour,
         yday = yday(pickup_datetime),
         work = as.numeric((hour %in% seq(8,18)) & (wday %in% c("Mon","Tue","Wed","Thu","Fri"))))
```

## Removing outliers

Removing trips that taking more than one day or less than 10 seconds.

```{r}
cb_dt = cb_dt %>%
  filter((dset == "test") | (duration < 24*3600 & duration > 10))
```

```{r}
# make sure we still have the same test set.
table(cb_dt$dset)
c(nrow(train_dt), nrow(test_dt))
```

## Save to files

```{r}
# for evaluation
set.seed(SEED)

train_dt = cb_dt %>% filter(dset == "train") %>% sample_n(100000) # will delete sample_n after testing the python code
smp_size = floor(0.75 * nrow(train_dt))
train_ind = sample(seq_len(nrow(train_dt)), size = smp_size)

exclude_col = c("dset")

fwrite(train_dt[train_ind, ] %>% select(-one_of(exclude_col)), file = "../data/processed/train_eval.csv")
fwrite(train_dt[-train_ind, ] %>% select(-one_of(exclude_col)), file = "../data/processed/test_eval.csv")
```


```{r}
fwrite(cb_dt %>% filter(dset == "train") %>% select(-one_of(exclude_col)), file = "../data/processed/train.csv")
fwrite(cb_dt %>% filter(dset == "test") %>% select(-one_of(c(exclude_col, "duration"))), file = "../data/processed/test.csv")
```

```{r}
cb_dt %>% filter(dset == "train") %>% summarise(mean(duration)/60)
cb_dt %>% filter(dset == "train") %>% summarise(median(duration)/60)

```


```{r}
cb_dt %>% filter(dset == "train") %>% mutate(mse = (duration - mean(duration))**2) %>% summarise(sum(mse)^0.5)
```
