---
title: "R Notebook"
output: html_notebook
---

This is for generating all shortest path for each trip.

# Load libraries

```{r}
library("data.table")
library("tidyverse")
library("ggplot2")
library("osrm")
library("profvis")
```

## Load data

```{r}
train_dt = fread("../data/raw/train.csv")
test_dt = fread("../data/raw/test.csv")
```

```{r}
cb_dt <- bind_rows(train_dt %>% mutate(dset = "train"),
                   test_dt %>% mutate(dset = "test",
                                     duration = NA))
cb_dt <- cb_dt %>% mutate(dset = factor(dset), id = row_number())
```

```{r}
# get trip duration and distance
## cb_dt
#fast_trip <-
cb_dt$fast_duration = 0
cb_dt$fast_distance = 0

#for (i in 1:nrow(cb_dt)) {
profvis(
  for (i in 1:1000) {
    cval = osrmRoute(src = cb_dt[i, c("row_id", "start_lng","start_lat")], cb_dt[i, c("row_id", "end_lng","end_lat")], overview = FALSE)
    cb_dt[i, "fast_duration"] = cval[["duration"]]
    cb_dt[i, "fast_distance"] = cval[["distance"]]
  }
)
```

```{r}
cb_dt[1:1000,] %>% ggplot(aes(x = duration, y = fast_duration * 60)) + geom_point() + geom_abline(slope = 1, intercept = 0)
```


The `duration` and `distance` features got from OSRM is very promising and can serve as great features for trip duration prediction. However, for 1000 samples, the program will take ~3 mins to run on my laptop and my home Internet. The total running time for all samples will be more than 24 hours.

After profiling the code, I found that the main bottleneck here is the http request to OSRM service (the `RCurl` call). If I have more time, I would write a distributed code to send request using multiple threads and deploy to several small size AWS (Amazon Web Service) instances. I would guess OSRM service will have some limitation for each IP address, so I will also need to tune the number of request/s per IP.

Data from OSRM provide a much more accurate distance and duration estimation between start and end location, if no other traffic condition. However, for a real-time service, we need to have our own service for fast route calculation.